The low-frequency Murchison radio environment:\\interference analysis and mitigation

### 2014-10-10 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Todays work

\begin{table*}
\caption{List of observations used in the analyses. The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (which includes false positives). The DVB column displays the fraction of visibilities that were unusable because of the presence of DVB signals, and which the flagger does not flag. Not all observations cover the DVB frequencies. }\label{tbl:obs-list}
\begin{center}%
\begin{tabular}{|l|l|r|r|r|c|}
\hline
\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DVB \\
\hline
EoR high& 2013-08-23 & 167.0--197.7              & 3h & 0.53\% & 0\% \\
GLEAM & 2013-08-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.94\% & 0\% \\
EoR low& 2013-08-26 & 138.9--169.6              & 6h & 0.81\% & ---\\
GLEAM & 2013-11-05 & 72.3--133.8, 138.9--230.8 & 7h & 1.27\% & 0\% \\
GLEAM & 2013-11-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.69\% & 0\% \\
EoR high& 2014-02-05 & 167.0--197.7              & 6h & 0.54\% & 0\%\\
GLEAM & 2014-03-16 & 72.3--133.8, 138.9--230.8 & 7h & 0.79\% & 0\% \\
GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 30\% \\
EoR high& 2014-04-10 & 167.0--197.7              & 6h & 0.68\% & 0\%\\
GLEAM & 2014-06-18 & 72.3--133.8, 138.9--230.8 & 7h & 0.98\% & 0\% \\
Orbcomm RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
\hline
\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.96\%} & \textbf{3\%}\\
\hline
\end{tabular}
\end{center}
\end{table*}

We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall). The observations are listen in Table~\ref{tbl:obs-list}.

@@

The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger}. This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band observation shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%.

@@

The residual noise levels after flagging can be used to validate whether the flagged data is free of RFI. In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations. Observation `GLEAM 2014-03-17' shows residual DVB interference, and it is clear that this RFI has not been adequately flagged. Therefore, observations have to be cleaned of DVB with another method. Some channels in the FM and 2-m amateur bands show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and RFI leakage. The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the large loss of data and residual RFI. Other than that, the RFI leakage is minimal and the RFI mitigation employed in \textsc{cotter} is sufficient to allow calibration and imaging without further effort. This includes the FM bands. This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.

\noindent\begin{figure*}%
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=8cm]{img/dvb_example}\includegraphics[width=8cm]{img/2014-03-17-dvb-stddevs}
\end{center}
\caption{RFI from Digital Video Broadcasting (DVB), which is detected in 1 of the 9 observations (GLEAM~2014-03-17) that cover the DVB frequencies. Left plot: dynamic spectrum for the worst affected snapshot; right plot: visibility RMS within each of the DVB bands for the affected night after flagging. Because of its broadband nature, this kind of RFI is not well flagged in the initial \textsc{aoflagger} step, but it is detectable in the global statistics. When observing at 182~MHz, this type of RFI makes most of the 30.72~MHz bandwidth unusable when it is visible. }
\label{fig:dvb}
\end{figure*}

@@

Despite not being detecting by the flagger, the right plot of Fig.~\ref{fig:dvb} shows that it is easy to recognize DVB interference from the data statistics, for example by testing whether the RMS in any of the occupied DVB bands exceeds the RMS in DVB band 4, which is not used for broadcasting. When DVB is detected in a snapshot, in most cases the best action will be to remove the entire snapshot, as it is not worth having special mechanisms in place to salvage the small amount of residual good data within an affected snapshot. With 2 out of 7 hours affected in 1 out of 9 observed nights, a rough probability of 3\% can be given for observing DVB, as is detailed in the last column of Table~\ref{tbl:obs-list}.

\noindent\begin{figure*}
\subcaptionbox{RFI contamination found in the 2 m amateur band (146 MHz).}[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/amateur-2m-band-example}\includegraphics[width=9.1cm]{img/amateur-2m-band-flagged}
\label{fig:amateur-2m}
}\\\vspace{3mm}%
\subcaptionbox{Short RFI burst of 2 s centred on 150.17 MHz. This shows that flagging at high time and frequency resolution is necessary to accurately detect RFI.}[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/150_2_mhz_example}\includegraphics[width=9.1cm]{img/150_2_mhz_flagged}
\label{fig:150_2}
}\\\vspace{3mm}%
\subcaptionbox{A weakly-observed transmitter centred on 156.66 MHz. This same frequency is flagged in several correlations, and it can therefore be assumed this is real RFI. This is an example in which the \textsc{aoflagger} detects an event that would not be detected by regular thresholding. }[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}
\label{fig:156_7}
}%
\caption{Examples of RFI events found in the EoR low band (138.9--197.7 MHz) in a 2-min snapshot with relatively high RFI contamination. These panels show the Stokes I amplitudes. In the right figures, the result of RFI detection is shown with purple. The horizontally flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
\end{figure*}

@@

\begin{itemize}
 \item An example of RFI found (e.g., the EoR DVB, ..)
 \item Performance
 \item Log N log S Histograms
\end{itemize}

\section{Discussion}
\begin{itemize}
 \item SKA -- how to implement a Cotter for SKA?
\end{itemize}

>>>>

\begin{table*}
\caption{List of observations used in the analyses. The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (which includes false positives). The DVB column displays the fraction of visibilities that were unusable because of the presence of DVB signals, and which the flagger does not flag. Not all observations cover the DVB frequencies. }\label{tbl:obs-list}
\begin{center}%
\begin{tabular}{|l|l|r|r|r|c|}
\hline
\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DVB \\
\hline
EoR high& 2013-08-23 & 167.0--197.7              & 3h & 0.53\% & 0\% \\
GLEAM & 2013-08-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.94\% & 0\% \\
EoR low& 2013-08-26 & 138.9--169.6              & 6h & 0.81\% & ---\\
GLEAM & 2013-11-05 & 72.3--133.8, 138.9--230.8 & 7h & 1.27\% & 0\% \\
GLEAM & 2013-11-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.69\% & 0\% \\
EoR high& 2014-02-05 & 167.0--197.7              & 6h & 0.54\% & 0\%\\
GLEAM & 2014-03-16 & 72.3--133.8, 138.9--230.8 & 7h & 0.79\% & 0\% \\
GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 30\% \\
EoR high& 2014-04-10 & 167.0--197.7              & 6h & 0.68\% & 0\%\\
GLEAM & 2014-06-18 & 72.3--133.8, 138.9--230.8 & 7h & 0.98\% & 0\% \\
Orbcomm RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
\hline
\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.96\%} & \textbf{3\%}\\
\hline
\multicolumn{4}{|r|}{\textbf{With uniform channel weighting}} & \textbf{1.13\%} & \\
\hline
\end{tabular}
\end{center}
\end{table*}

We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall). The observation nights are listed in Table~\ref{tbl:obs-list}.

@@

The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger}. This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band observation shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global expected RFI ratio in the 72.3--230.8 MHz range.

@@

The residual noise levels after flagging can be used to validate whether the flagged data is free of RFI. In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations. Observation `GLEAM 2014-03-17' shows residual DVB interference, and it is clear that this RFI has not been adequately flagged. Therefore, observations have to be cleaned of DVB with another method. Additionally, some channels in the FM and 2-m amateur bands show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and because of RFI leakage, but because the effect is small these frequencies can be calibrated and imaged without a problem. The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI. Other than that, the RFI leakage is minimal. With exception of the ORBCOMM frequencies and DVB affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts. This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.

\noindent\begin{figure*}%
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=10cm]{img/dvb_example}\includegraphics[width=5cm]{img/2014-03-17-dvb-stddevs}
\end{center}
\caption{RFI from Digital Video Broadcasting (DVB), which is visible in 1 of the 9 observations (GLEAM~2014-03-17) that cover the DVB frequencies. Left plot: dynamic spectrum for the worst affected snapshot; right plot: visibility RMS in a few of the DVB bands for the affected night after flagging. Because of its broadband nature, this kind of RFI is not well flagged in the initial \textsc{aoflagger} step, but it is detectable in the global statistics. When observing at 182~MHz, this type of RFI makes most of the 30.72~MHz bandwidth unusable when it is visible. }
\label{fig:dvb}
\end{figure*}

@@

Although the flagger does not adequately flag DVB, the right plot of Fig.~\ref{fig:dvb} shows that it is easy to recognize DVB interference from the data statistics. An option is to test whether the visibility RMS of a snapshot in any of the occupied DVB bands exceeds the RMS in DVB band 4, which is not used for broadcasting. When DVB is detected in a snapshot, in most cases the best action will be to remove the entire snapshot, as it is not worth having special mechanisms in place to salvage the small amount of residual good data within an affected snapshot. Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability of observing DVB is roughly 3\%. This is detailed in the last column of Table~\ref{tbl:obs-list}.

\noindent\begin{figure*}
\subcaptionbox{RFI contamination found in the 2 m amateur band (146 MHz).}[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/amateur-2m-band-example}\includegraphics[width=9.1cm]{img/amateur-2m-band-flagged}
\label{fig:amateur-2m}
}\\\vspace{3mm}%
\subcaptionbox{Short RFI burst of 2 s centred on 150.17 MHz. Detection of this kind of RFI requires flagging at high time and frequency resolution.}[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/150_2_mhz_example}\includegraphics[width=9.1cm]{img/150_2_mhz_flagged}
\label{fig:150_2}
}\\\vspace{3mm}%
\subcaptionbox{A weakly-observed transmitter centred on 156.66 MHz. This same frequency is flagged in several correlations, and it can therefore be assumed this is real RFI. This is an example in which the \textsc{aoflagger} detects an event that would not be detected by regular thresholding. }[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}
\label{fig:156_7}
}%
\caption{Examples of RFI events found in the EoR low band (138.9--169.6 MHz) in a 2-min snapshot with relatively high RFI contamination. These panels show the Stokes I amplitudes. In the right figures, the result of RFI detection is shown with purple. The horizontally flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
\end{figure*}

\noindent\begin{figure*}%
% obs data: 1075641128
\begin{center}\hspace*{-0.2cm}\includegraphics[width=8cm]{img/EoR-high-band-DVB-burst-example-noflags}\includegraphics[width=8cm]{img/EoR-high-band-DVB-burst-example-withflags}
\end{center}
\caption{???. }
\label{fig:dvb-burst}
\end{figure*}

@@

\subsection{Computational performance of \textsc{cotter}}
Because \textsc{cotter} processes the data at the highest resolution, its computational performance is an important consideration. Because of the large volume of MWA data, its runtime is strongly dependent on the input-output (IO) performance of the host system. Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent. In \citet{lofar-radio-environment} it is shown that the performance of the \textsc{aoflagger} is in normal situations dominated by IO, and hence the same holds for \textsc{cotter}: on a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup, processing a single 2~min snapshot with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, approximately 3 min are spend on reading the data, 5 min on RFI detection and 3.5 min on writing the data. Real-time processing can therefore be achieved by using 6 of such nodes in parallel.

\begin{itemize}
 \item Describe RFI examples (EoR low band \& high band)
 \item Performance
 \item Log N log S Histograms
\end{itemize}

\section{Discussion}
\begin{itemize}
 \item SKA -- how to implement a Cotter for SKA?
 \item PAPER etc needs to use combinatorial/morphological flagging
\end{itemize}
\subsection{Comparison with LOFAR}
